---
title: 'PHB 222 Final Project: Initial Data Exploration'
author: "Himani Yalamaddi"
date: "2025-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

library(readxl) # read in excel data (facilities info)
library(janitor) # to clean up variable names automatically

# api packages to get census boundaries from fcc area api
library(httr)
library(jsonlite)

# to experiment with using census api 
# (optional -- i'm having trouble getting an census API key)
# library(censusapi)
``` 

## background 

our articles of interest: 

1. charleston, sc: https://ajph.aphapublications.org/doi/full/10.2105/AJPH.2012.300700
2. atlanta, ga: https://www.mdpi.com/1660-4601/13/8/747 

## data sources / import all data 

**data sources**: 

1. tri basic plus data (2023, file #1a *tri-basic-plus-1a-2023-sd.csv*): https://www.epa.gov/toxics-release-inventory-tri-program/tri-basic-plus-data-files-calendar-years-1987-present 

2. 2020 dicennial census demographics (*dem-dicennial-2020-sd.csv*): this contains race/ethnicity, gender, age, housing units, household information.

3. 2023 acs (american community survey) income/poverty (*income-acs2023.csv*): this contains median/mean income by households. even though census data is considered more accurate, especially at the tract level, income is only collected through the acs. (https://data.census.gov/table/ACSST1Y2023.S1901?q=Income%20and%20Poverty)

4. 2023 acs (american community survey) educational attainment (*education-acs2023-sd.csv*): this file contains educational attainment or all census tract members ages 25+; male and female are distinct here (atlanta paper).

5. fcc area and census conversion API: https://geo.fcc.gov/api/census/. i used this tutorial: https://www.dataquest.io/blog/r-api-tutorial/ 


*note about census blocks: "block_fips" is 15 digits b/c it includes an additional 4-digit "block" (more detailed) code. we only want the first 11 digits, which identify census tracts, bc census blocks may be too small of a unit to be useful for us (while census tracts usually aim to be populated in some way, many census blocks have no population whatsoever, which isn't helpful for us). in addition, our demographic info uses 11-digit census tract IDs, so we'll use tracts as our unit of measurement.*


### get one-row-one-emission data (75 emitting facilities) 

this file contains a row for each of the *emissions* reported by tri facilities in 2023. note that one facility can emit more than one chemical. below are a few notable facilities in SD in terms of # of unique chemicals released: 

1. 92113RCSND2295E: TESORO - SAN DIEGO TERMINAL	(MARATHON PETROLEUM CORP): 9 chemicals
2. 92055MRNCRPOBOX: US MARINE CORPS MCB CAMP PENDLETON: 7 chemicals
2. 92121FRZPN6625M: SHERWIN-WILLIAMS CO (6 chemicals) 

facilities can report using either one of two forms (variable `x1_form_type`). form A is more simplistic, and the more elaborate form R is only needed if: 

1. the toxic chemical is not a PBT chemical 
2. the toxic chemical was not manufactured, processed, or otherwise used in excess of 1,000,000 pounds 
3. the total annual amount released of the toxic chemical did not exceed 500 pounds. 

link to emission-level variable documentation: https://www.epa.gov/system/files/documents/2024-09/file_type_1a_1.pdf


```{r, warning=FALSE, message = FALSE}
# 0. creation of tri-basic-plus-1a-2023-sd.csv from US_1a_2023.xlsx
# read in toxics release inventory emissions data 
# (file 1a downloaded from basic plus files):
# emissions <- read_excel("data/US_1a_2023.xlsx")
# emissions_SD <- emissions %>% filter(`14. FACILITY STATE` == "CA" & 
                                      #`13. FACILITY COUNTY` == "SAN DIEGO")
# write_csv(emissions_SD, "data/tri-basic-plus-1a-2023-sd.csv")

emissions_SD <- read_csv("data/tri-basic-plus-1a-2023-sd.csv")

# clean var names: replace spaces w/ "_"
# leading x: b/c R doesn't like variable names to begin w/ numbers
emissions_SD <- clean_names(emissions_SD)

# let's isolate the names of key variables that capture emissions for future use:
# below are some findings (and why i removed some variables)
# all 2023
# no trade secrets
# all facilities in SD county/CA
# no stream B emissions + beyond
# no x206_total_surface_impoundments
# no x194_total_on_site_rcra_subtitle_c_landfills_pounds
# no x198_total_other_on_site_landfills
# no x190_total_on_site_landfills
# no x187_total_on_site_underground_injection

emissions_SD <- emissions_SD %>% select(
                        # facility information:
                        x9_trifd, 
                        x1_form_type,
                        x10_facility_name,
                        x71_parent_company_name,
                        x8_date_signed,
                        x47_latitude,
                        x48_longitude,
                        # chemical/emission information:
                        x90_revision_code_1,
                        x81_chemical_name,
                        x84_classification, 
                        x85_unit_of_measure,
                        x83_elemental_metal_included,
                        x86_hazardous_air_pollutant_haps,
                        x87_carcinogen,
                        x88_pfas_ind,
                        x89_metal_ind,
                        x109_maximum_amount_on_site, 
                        x112_total_fugitive_air_emissions, 
                        x116_total_stack_air_emissions, 
                        x118_total_air_emissions, # 112 + 116
                        x119_discharges_to_stream_a_stream_name, 
                        x122_total_discharges_to_stream_a,
                        x214_total_other_surface_impoundments, # one facility
                        x218_total_other_disposal, 
                        x220_total_on_site_land_releases, # one facility
                        x246_total_transferred_off_site_for_disposal,
                        x252_total_transferred_off_site_for_recycling, 
                        x255_total_transferred_off_site_for_energy_recovery, 
                        x263_total_transferred_off_site_for_treatment,
                        x265_total_transferred_off_site_for_further_waste_management,
                        x266_total_potw_transfer, # potw = publicly owned treatment works
                        x263_total_transferred_off_site_for_treatment,
                        x269_treated_on_site_current_year,
                        x270_total_on_site_waste_management, 
                        # MORE facility info, kept in case we wanted to create an interactive map:
                        x11_facility_street, x12_facility_city,
                        x15_facility_zip_code, x17_tribe_name, 
                        x24_entire_facility_ind, 
                        x25_partial_facility_ind,
                        x26_federal_facility_ind,
                        x27_goco_facility_ind,
                        x41_primary_naics_code,
                        x31_public_contact_phone, 
                        x33_public_contact_email)
``` 

### identify "superpolluters" -- FUTURE WORK 

```{r}
# check dimensions of new emission-level data set: 
dim(emissions_SD) # 154 emissions, 44 variables about each

# explore how many unique chemicals each of the facilities reported in 2023
table(emissions_SD$x9_trifd) # 75 total facilities
```

### create facility-level dataset to merge with census tract data 

below, we create a new data set with facility information and merge it with census tract boundaries and tract-level demographics information:


```{r}
# 1. create a new table JUST for facility information 
facilities_SD <- emissions_SD %>% select(x9_trifd, 
                        #x1_form_type,
                        x90_revision_code_1,
                        x10_facility_name,
                        x71_parent_company_name,
                        x8_date_signed,
                        x47_latitude,
                        x48_longitude,
                        x11_facility_street, x12_facility_city,
                        x15_facility_zip_code, x17_tribe_name, 
                        x24_entire_facility_ind, 
                        x25_partial_facility_ind,
                        x26_federal_facility_ind,
                        x27_goco_facility_ind,
                        x41_primary_naics_code,
                        x31_public_contact_phone, 
                        x33_public_contact_email) %>% 
  group_by(x9_trifd) %>% filter(x8_date_signed == max(x8_date_signed)) %>%
  ungroup() %>% distinct()

# 2. prelim cleaning up of facilities-level data: 
# there's one facility with everything duplicated except for contact info: "91950HNSNG61W12"
# we'll delete one version of this at random 
# --? FUTURE WORK; NOT NEEDED FOR CENSUS TRACT MERGING
# emissions_SD %>% filter(x9_trifd == "91950HNSNG61W12")

# emission recalculations/repeats 
# (will deal with this later; for now i didn't care to clean this up)
table(emissions_SD$x90_revision_code_1) 
# 3 facilities reported RR4 (recalculations)
# 1 facility reported RR5 (other reasons)

# isolate information for some key facilities to explore
# this one releases 9 unique chemicals!:
emissions_SD %>% filter(x9_trifd == "92113RCSND2295E")
``` 

### get census tract/demographics information, merge, and clean 

1. 2020 dicennial census demographics (*dem-dicennial-2020-sd.csv*): this contains race/ethnicity, gender, age, housing units, household information.

2. 2023 acs (american community survey) income/poverty (*income-acs2023.csv*): this contains median/mean income by households. even though census data is considered more accurate, especially at the tract level, income is only collected through the acs. (https://data.census.gov/table/ACSST1Y2023.S1901?q=Income%20and%20Poverty)

3. 2023 acs (american community survey) educational attainment (*education-acs2023-sd.csv*): this file contains educational attainment or all census tract members ages 25+; male and female are distinct here (atlanta paper).

**FUTURE WORK**: instead of designating a census tract as a "success" if the facility is in the tract, we could take a 1 mi buffer from each of the facilities. aka: "if a tri facility's 1mi buffer is within a census tract, then we include that tract as a census tract with a tri facility." i think this would help us get more "successes." 

below, we isolate a few (~30) variables in our topics of interest; not all will be used or be considered variables of interest! 


```{r}
# 1. read in all predictor data from census/acs data:
# 1.1. read in demographics (2020 dicennial census, census.gov) data
demographics <- read.csv("data/dem-dicennial-2020-sd.csv") 
dem <- demographics[-1, ] # remove california-wide stats 
dem <- clean_names(dem)
dem$geography <- substr(dem$geography, start = 10, stop = 20) 
# clean census tract ID for easy merging

dem <- dem %>% select(geography,
               # total population
               count_race_total_population,
               
               # race/ethnicity information
               percent_race_total_population_one_race_white,
               percent_hispanic_or_latino_by_race_total_population_hispanic_or_latino,
               percent_race_total_population_one_race_black_or_african_american, 
               percent_race_total_population_one_race_asian,
               percent_race_total_population_one_race_native_hawaiian_and_other_pacific_islander,
               percent_race_total_population_one_race_american_indian_and_alaska_native,
               percent_race_total_population_one_race_some_other_race,
               percent_race_total_population_two_or_more_races,
               percent_total_races_tallied_1_total_races_tallied_black_or_african_american_alone_or_in_combination_with_one_or_more_other_races, 
               percent_total_races_tallied_1_total_races_tallied_black_or_african_american_alone_or_in_combination_with_one_or_more_other_races, 
               percent_total_races_tallied_1_total_races_tallied_white_alone_or_in_combination_with_one_or_more_other_races,
               
               # housing information 
               percent_housing_occupancy_total_housing_units_vacant_housing_units, # (used in atlanta study)
               percent_housing_tenure_occupied_housing_units_renter_occupied_housing_units)


# 1.2. read in income (2023 ACS, census.gov) data
income <- read.csv("data/income-acs2023-sd.csv")
names(income) <- income[1,] # make row 1 header
inc <- income[-1, ] # remove first row with info
inc <- clean_names(inc)
inc$geography <- substr(inc$geography, start = 10, stop = 20)

inc <- inc %>% select(geography,
               # total # of households
               estimate_households_total,
               
               # income in dollars
               estimate_households_median_income_dollars,
               estimate_households_mean_income_dollars)


# 1.3 read in educational attainment (2023 ACS, census.gov) data
# this is a HUGE file (771 columns), MUST be pared down before merging
education <- read.csv("data/education-acs2023-sd.csv")
names(education) <- education[1,] # make row 1 header
edu <- education[-1, ] # remove first row with info
edu <- clean_names(edu)
edu$geography <- substr(edu$geography, start = 10, stop = 20)

edu <- edu %>% select(geography,
               # total population educational attainment
               estimate_percent_age_by_educational_attainment_population_25_years_and_over_bachelors_degree_or_higher, 
               estimate_percent_age_by_educational_attainment_population_25_years_and_over_high_school_graduate_or_higher, 
               estimate_percent_age_by_educational_attainment_population_25_years_and_over_graduate_or_professional_degree,
               estimate_percent_age_by_educational_attainment_population_25_years_and_over_less_than_9th_grade,
               
               # female population educational attainment
               estimate_percent_female_age_by_educational_attainment_population_25_years_and_over_bachelors_degree_or_higher, 
               estimate_percent_female_age_by_educational_attainment_population_25_years_and_over_graduate_or_professional_degree,
               estimate_percent_female_age_by_educational_attainment_population_25_years_and_over_high_school_graduate_or_higher,
               estimate_percent_female_age_by_educational_attainment_population_25_years_and_over_less_than_9th_grade,
               
               # male population educational attainment
               estimate_percent_male_age_by_educational_attainment_population_25_years_and_over_bachelors_degree_or_higher, 
               estimate_percent_male_age_by_educational_attainment_population_25_years_and_over_graduate_or_professional_degree,
               estimate_percent_male_age_by_educational_attainment_population_25_years_and_over_high_school_graduate_or_higher,
               estimate_percent_male_age_by_educational_attainment_population_25_years_and_over_less_than_9th_grade, 
               
               # male and female median earnings in the last 12 months (2023)
               estimate_female_median_earnings_in_the_past_12_months_in_2023_inflation_adjusted_dollars_population_25_years_and_over_with_earnings, 
               estimate_male_median_earnings_in_the_past_12_months_in_2023_inflation_adjusted_dollars_population_25_years_and_over_with_earnings)


# 2. get census tracts for each of our facilities using lat/long coords!
facility_tract_IDs <- vector("list", nrow(facilities_SD))
for (i in 1:nrow(facilities_SD)) {
  latitude <- facilities_SD$x47_latitude[i]
  longitude <- facilities_SD$x48_longitude[i]
  result <- GET("https://geo.fcc.gov/api/census/area", 
           query = list(lat = latitude, lon = longitude, 
                        censusYear = 2020, format="json"))
  facility_tract_IDs[[i]] <- fromJSON(rawToChar(result$content))[[2]][["block_fips"]]
}

# 3. extract all unique census tract ids from census block ids:
facility_tracts <- lapply(facility_tract_IDs, substr, start = 1, stop = 11)
# the total number of times each of the census tract names show up
# could be taken to mean the number of facilities in each tract -- BUT...
final_tracts <- table(unlist(facility_tracts))

# ...WARNING: there are three facilities are in 2 census tracts?: 
### lapply(facility_tract_IDs, unique)
# not 100% sure how this is possible -- will look into it
# for now, i included all tracts

# CHECK: make sure all of our facilities are in known tracts:
# all(unlist(facility_tracts) %in% dem$geography) # GOOD

# 4. create variables in demographics table for facility count in that tract
dem$facility_binary <- dem$geography %in% unlist(facility_tracts)
dem$facility_counts <- ifelse(dem$facility_binary, final_tracts[dem$geography], 0)

# 5. merge all data (dem, inc, edu) by the "geography" column; should be 737 census tracts
# all(dem$geography %in% inc$geography)
# all(dem$inc$geography %in% edu$geography)
# any(c(duplicated(dem$geography), duplicated(inc$geography), duplicated(edu$geography)))
dat <- dem %>% inner_join(inc) %>% inner_join(edu)

write.csv(dat, "merged_data_v1.csv")
``` 


## pre-preliminary look at the data: 

in total, it looks like we have around 40 "successes" out of 737 census tracts; this will likely limit the number of variables that we can include in our model to about 2 to satisfy the "rule of thumb" 15 successful observations per variable. **dr. messer said the way to deal with this, generally, is to start by running one-by-one individual regressions with each variable of interest first, before doing anything else.**

```{r}
# get success/failure (whether there's a facility in the tract) counts
table(dat$facility_binary)
40/nrow(dem) # proportion of successes: ~5.4% success rate -- pretty low

# get the # of counts in each "successful" tract: 
final_tracts # one tract with 13: wow!

# let's explore tract 06073008350 (tract 83.50), which has 13 facilities in it
# WARNING: this will print ALL AVAILABLE VARIABLES WHEN KNITTED!
dat %>% filter(geography == "06073008350")
# population: 6228
``` 





```{r}



``` 










***NO LONGER USING: FACILITIES-LEVEL DATA (data set # 4)***:

```{r, warning = F, echo = F}
### get facilities data (70 emitting facilities in San Diego) - NO LONGER USING !!!
# 1. toxics release inventory: facilities data (2023)
# facilities <- read_excel("data/US_4_2023_excel.xlsx")
# facilities_SD <- facilities %>% filter(`9. STATE` == "CA" & `8. COUNTY` == "SAN DIEGO")
# confirmed no duplicates + and no NA values in either latitude or longitude information

# understanding the relationship between form 4 and 1a: 
# unique(emissions_SD$`9. TRIFD`) %in% unique(facilities_SD4$`4. TRIFD`) # 5 "missing" in facilities_SD4; no NA
# unique(facilities_SD4$`4. TRIFD`) %in% unique(emissions_SD$`9. TRIFD`) # all included; no NA
# note that all reported facilities are in the emissions data;
# not all emitters are in our facilities data:

# emissions_SD$`9. TRIFD`[miss_ids <- which( !(unique(emissions_SD$`9. TRIFD`) %in% unique(facilities_SD$`4. TRIFD`) ))]
# emissions_SD$`10. FACILITY NAME`[miss_ids]

# emissions_SD[emissions_SD$`9. TRIFD` == "92020CHMTR1150W", ]
# "92020CHMTR1150W" %in% emissions_SD$`9. TRIFD`
``` 

