---
title: "models"
date: "2025-03-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
```

## get data 
```{r}
final_imputed <- read.csv("data/final-imputed.csv")
final_subset <- read.csv(("data/final_subsetted_not_imputed.csv"))
```

```{r}
library(table1)

# population
label(final_subset$total_pop) <- "Total Population"

# race/ethnicity
label(final_subset$people_of_color_pct) <- "% People of Color" # 100 - white only, non-hispanic
label(final_subset$hispanic_pct) <- "% Hispanic/Latino"
label(final_subset$black_only_pct) <- "% Black Only"
label(final_subset$asian_only_pct) <- "% Asian Only"

# socioeconomics
label(final_subset$median_income) <- "Median Household Income"
label(final_subset$renter_pct) <- "% of Housing Renter-Occupied"
label(final_subset$bachelors_pct ) <- "% Population with Bachelors degree or higher"
label(final_subset$female_bachelors_pct) <- "% Women with Bachelors degree or higher"

label(final_subset$less_than_9th_pct ) <- "% Population with 9th Grade or less"
label(final_subset$female_less_than_9th_pct) <- "% Women with 9th Grade or less"


render_median <- function(x, ...) sprintf("<b>%0.1f</b> <i>(%0.1f-%0.1f)</i>", 
                                          median(x, na.rm = T), 
                                          min(x, na.rm = T), 
                                          max(x, na.rm = T))


table1(~ total_pop + people_of_color_pct + hispanic_pct + black_only_pct + asian_only_pct
        + median_income + renter_pct + 
         bachelors_pct + female_bachelors_pct + less_than_9th_pct + female_less_than_9th_pct | host, 
       data = final_subset,
       overall = FALSE,
       render.continuous = render_median)

```


## pre-preliminary look at the data: 

in total, it looks like we have around 40 "successes" out of 737 census tracts; this will likely limit the number of variables that we can include in our model to about 2 to satisfy the "rule of thumb" 15 successful observations per variable. **dr. messer said the way to deal with this, generally, is to start by running one-by-one individual regressions with each variable of interest first, before doing anything else.**

```{r}
# VARIABLES OF INTEREST: 
# 1. "affected": binary variable (host vs. non-host = outside of 3km limit)
# 2. "facility_buffer_counts": counts data for ZIP model

# BINARY (affected)
table(final_subset$host)
# 487 non-hosts, 249 hosts
# 249 / (487 + 249) = 0.3383152

# COUNTS (facility_buffer_counts)
# paper: 0 vs 1 vs 1+ was how the paper did counts analysis 
# here we DO NOT separate into host/non-host.
# if the buffer is in the tract at all, then we add it
table(final_subset$facility_counts) # ALL
#   0   1   2   3   4   5   6   7   8   9  10  11  17  24 
# 357 124  74  62  27  34  27  15   6   4   3   1   1   1 

table(ifelse(as.numeric(final_subset$host) == 2, final_subset$facility_counts, 0)) # ONLY HOSTS
#   0   1   2   3   4   5   6   7   8   9  10  17  24 
# 487  44  52  47  22  31  23  15   6   4   3   1   1 


# let's explore tract 06073008350 (tract 83.50), which has 13 facilities in it
# WARNING: this will print ALL AVAILABLE VARIABLES WHEN KNITTED!
final_subset %>% filter(geo == "06073008350")
# population: 6228

### RENTER-OCCUPIED
mu <- plyr::ddply(final_subset, "host", summarise, 
            grp.mean = mean(as.numeric(renter_pct)))

ggplot(final_subset, aes(x = as.numeric(renter_pct), 
                fill = factor(host), 
                   color = factor(host))) + 
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + 
  geom_vline(data = mu, aes(xintercept = grp.mean, 
                        color = factor(host)), 
             linetype = "dashed") +
  theme(legend.position = "top") + 
  labs(title = "Distribution of % of renter-occupied units, by facility presence") + 
  xlab("% of renter-occupied units") + 
  ylab("Count")

### % POC
mu <- plyr::ddply(final_subset, "host", summarise, 
            grp.mean = mean(as.numeric(people_of_color_pct)))

ggplot(final_subset, aes(x = as.numeric(people_of_color_pct), 
                fill = factor(host), 
                   color = factor(host))) + 
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + 
  geom_vline(data = mu, aes(xintercept = grp.mean, 
                        color = factor(host)), 
             linetype = "dashed") +
  theme(legend.position = "top") + 
  labs(title = "Distribution of % of POC, by facility presence") + 
  xlab("% of POC") + 
  ylab("Count")


### MEDIAN INCOME
# mu <- plyr::ddply(final, "host", summarise, 
#             grp.mean = mean(as.numeric(median_income), 
#             na.rm = T))
# 
# ggplot(final_subset, aes(x = as.numeric(median_income), 
#                 fill = factor(host), 
#                    color = factor(host))) + 
#   geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + 
#   #geom_vline(data = mu, aes(xintercept = grp.mean, 
#                         #color = factor(affected)), 
#              #linetype = "dashed") +
#   theme(legend.position = "top") + 
#   labs(title = "Distribution of median household income, by facility presence") + 
#   xlab("Median household income") + 
#   ylab("Count")
  

### EDUCATION
mu <- plyr::ddply(final, "host", summarise, 
            grp.mean = mean(as.numeric(bachelors_pct), 
            na.rm = T))

ggplot(final_subset, aes(x = as.numeric(bachelors_pct), 
                fill = factor(host), 
                   color = factor(host))) + 
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + 
  geom_vline(data = mu, aes(xintercept = grp.mean, 
                        color = factor(host)), 
             linetype = "dashed") +
  theme(legend.position = "top") + 
  labs(title = "Distribution of % w/ bachelor's degree, by facility presence") + 
  xlab("Median household income") + 
  ylab("Count")

as.numeric(dat$estimate_households_median_income_dollars)

# pink: distribution of age for non-opioid users
# blue: distribution of age for opioid users
``` 


## Logistic regression model 

```{r}
final_imputed$host <- as.factor(final_imputed$host)
final_imputed$host <- relevel(final_imputed$host, ref = "Non-Host")

summary(m1 <- glm(host ~
                   # people_of_color_pct +
                    #hispanic_pct +
                    #black_only_pct +
                    #asian_only_pct +
                    median_income_scaled + 
                    renter_pct +
                    vacant_housing_pct +
                   # bachelors_pct + 
                    high_school_pct, 
                    # female_bachelors_pct,
             family = "binomial", 
             data = final_imputed))


m2_back <- step(m1, direction = "backward")
summary(m2_back)


summary(m11 <- glm(host ~
                   # people_of_color_pct +
                    #hispanic_pct +
                    black_only_pct + 
                    asian_only_pct+
                    # white_only_non_hispanic_pct + 
                    renter_pct + 
                    # median_income +
                    # bachelors_pct,
                    # high_school_pct, 
                    female_bachelors_pct,
             family = "binomial", 
             data = final_imputed[-c(116, 281), ])) 



summary(m11 <- glm(host ~
                   # people_of_color_pct +
                    hispanic_pct +
                    black_only_pct + 
                    asian_only_pct +
                    # white_only_non_hispanic_pct + 
                    renter_pct + 
                    # median_income +
                    # bachelors_pct,
                    # high_school_pct, 
                    female_bachelors_pct,
             family = "binomial", 
             data = final_imputed)) 
vif(m11)
generalhoslem::logitgof(m11$y, fitted(m11))
plot(m11, 4) # 116, 281


# bonferroni correction
# Define the predictors (all columns except response variable)
predictors <- setdiff(names(final_imputed), 
                      c("host", "geo", "facility_counts", "hosts_facility_counts", "total_pop"))

# Initialize a data frame to store significant predictors
significant_predictors <- data.frame(Predictor = character(), P_Value = numeric(), stringsAsFactors = FALSE)

# Loop through each predictor to perform pairwise logistic regression
for (predictor in predictors) {
  # Fit logistic regression model
  model <- glm(final_imputed$host ~ final_imputed[[predictor]], family = binomial)
  
  # Extract p-value
  p_value <- summary(model)$coefficients[2, 4]  # Second row, fourth column (p-value for predictor)
  
  # Store only significant predictors (p < 0.3)
  if (!is.na(p_value) && p_value < (0.05 / length(predictors))) {
    significant_predictors <- rbind(significant_predictors, data.frame(Predictor = predictor, P_Value = p_value))
  }
}

# Sort results by p-value
significant_predictors <- significant_predictors[order(significant_predictors$P_Value), ]

# Print significant predictors
print(significant_predictors)

m1_back <- step(m1)
summary(m1_back)
exp(confint(m1_back))

# multicollinearity
vif(m1_back)


# model after pairwise regression: 
summary(m3 <- glm(host ~
                    renter_pct + 
                    median_income + 
                    high_school_pct + 
                    black_only_pct + 
                    asian_only_pct,
             family = "binomial", 
             data = final_imputed))

AIC(m3)
AIC(m1_back)

cor(final_imputed$high_school_pct, final_imputed$median_income)

vif(m3)


library(generalhoslem)

generalhoslem::logitgof(m1_back$y, fitted(m1_back))
generalhoslem::logitgof(m3$y, fitted(m3))
``` 


## Zero-inflated model 

Here, we're looking at the **counts** of buffers that make up in each of the host tracts. In doing 

```{r}
mean(final_subset$facility_counts[final_subset$facility_counts> 0])
var(final_subset$facility_counts[final_subset$facility_counts > 0]) # overdispersed

# how are these 12 NAs different than the rest of the population?
sum(is.na(as.numeric(final$estimate_households_median_income_dollars)))

# for some reason, "system is computationally singular" when our median income variable is included. why!!!




library(pscl)
summary(zi1 <- pscl::zeroinfl(hosts_facility_counts ~
                   # median_income_scaled + 
                    renter_pct +
                    hispanic_pct +
                    black_only_pct +
                   asian_only_pct + 
                    female_bachelors_pct |
                   #median_income_scaled + 
                   renter_pct +
                   hispanic_pct +
                    black_only_pct +
                    asian_only_pct + 
                    female_bachelors_pct,
                      dist = "poisson",
                      data = final_subset))
```